---
title: "FinalProject"
author: "Elizabeth Lattanzi"
date: "2025-11-30"
output: html_document
---

```{r setup, include=FALSE}
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(brms)
library(loo)
library(janitor)
library(stringr)
library(lubridate)
library(purrr)
library(bayesplot)
library(Metrics)
```

### Load and clean wildfire data

```{r}

# Aggregate burned acreage data and group by county ID and year, and add a join

path <- "fire24_1.gdb"
fire_data <- st_read(path, layer = "firep24_1") %>%
  st_drop_geometry()
county_year_burned <- fire_data %>%
  group_by(UNIT_ID, YEAR_) %>%
  summarise(total_burned_acres = sum(GIS_ACRES, na.rm = TRUE), .groups = "drop")


unit_lookup <- tibble::tibble(
  UNIT_ID = c("BTU", "LNU", "SHF", "SCU", "MNF", "TGU", "NEU"),
  county = c("Butte County", "Napa County", "Shasta County", "Santa Clara County", 
             "Mendocino County", "Tehama County", "Nevada County"))

county_year_named <- county_year_burned %>%
  left_join(unit_lookup, by = "UNIT_ID")


```

### Load and clean lung cancer data

```{r}


# Clean and standardize data, filter out rows where key variable is missing  

lung_data_raw <- read_csv("incd.csv", skip = 8)
names(lung_data_raw) <- make.names(names(lung_data_raw))
lung_data <- lung_data_raw %>%
  clean_names() %>%
  filter(!is.na(age_adjusted_incidence_rate_rate_note_cases_per_100_000)) %>%
  mutate(county = gsub("\\s*\\(.*\\)", "", county))


```

### Merge and derive exposure variables
```{r}

#  Join the cleaned lung cancer data with the wildfire exposure data, using the county name as the key, merge into the final clean dataset

merged_data <- lung_data %>%
  left_join(county_year_named, by = "county") %>%
  mutate(age_adjusted_rate = as.numeric(age_adjusted_incidence_rate_rate_note_cases_per_100_000), incidence_numeric = age_adjusted_rate)
```

### Merge in PM2.5 data and preprocess
```{r}

# Add in the PM2.5 dataset, merge into the final dataset
# Z-score standardize the pm25_avg, log-transform total burned acres 

pm_data <- "pm2.5_data"
pm_data <- list.files(pm_data, pattern = "\\.csv$", full.names = TRUE)

pm25_all <- map_dfr(pm_data, function(file) {
  read_csv(file, show_col_types = FALSE) %>%
    filter(`State Code` == "06", `Parameter Name` == "PM2.5 - Local Conditions") %>%
    mutate(date = as.Date(`Date Local`), year = year(date), fips = paste0(`State Code`, str_pad(`County Code`, 3, pad = "0"))) %>%
    group_by(fips, year) %>% summarise(pm25_avg = mean(`Arithmetic Mean`, na.rm = TRUE), .groups = "drop")})

merged_data_pm25 <- merged_data %>%
  mutate(fips = as.character(fips), YEAR_ = as.numeric(YEAR_)) %>%
  left_join(pm25_all, by = c("fips" = "fips", "YEAR_" = "year")) %>%
  rename(urban_rural = x2023_rural_urban_continuum_codes_rural_urban_note) %>%
  mutate(pm25_avg_z = scale(pm25_avg, center = TRUE, scale = TRUE),log_burned_acres = log1p(total_burned_acres))

```

### Fit Bayesian model
```{r}

# Set the Normal(0, 5) prior on all regression coefficients, use a student_t(3, 0, 10) prior on the model's residual standard deviation 


bayes_model_pm25 <- brm(
  formula = incidence_numeric ~ pm25_avg_z * log_burned_acres + urban_rural,
  data = merged_data_pm25,
  family = gaussian(),
  prior = c(set_prior("normal(2, 3)", class = "b"), set_prior("student_t(3, 0, 10)", class = "sigma")),
  seed = 15,
  chains = 5,
  iter = 5000,
  warmup = 1000,
  control = list(adapt_delta = 0.95))



```

### Evaluate model

```{r}

# print summary statistics, posterior predictive checks 

summary(bayes_model_pm25)
pp_check(bayes_model_pm25)


```

The posterior predictive checks suggests that there are two distinct clusters with different incidence rates. This suggests there could be an interaction between terms. 




### Posterior prediction

```{r}

# generate posterior predictions, extract the estimate, and the 95% interval 


bayes_data_pred <- merged_data_pm25 %>%
  filter(!is.na(log_burned_acres), !is.na(urban_rural), !is.na(incidence_numeric), !is.na(pm25_avg_z))

fitted_values <- fitted(bayes_model_pm25, newdata = bayes_data_pred)
bayes_data_pred$predicted_mean <- fitted_values[, "Estimate"]
bayes_data_pred$predicted_lower <- fitted_values[, "Q2.5"]
bayes_data_pred$predicted_upper <- fitted_values[, "Q97.5"]


```

### Print Summary Statistics
```{r}

# compute residuals, MAE, R^2, print 95% interval, plot prediction accuracy 

bayes_data_pred$residuals <- bayes_data_pred$incidence_numeric - bayes_data_pred$predicted_mean
mae_val <- mae(bayes_data_pred$incidence_numeric, bayes_data_pred$predicted_mean)
r2_val <- bayes_R2(bayes_model_pm25)
r2_summary <- summary(r2_val)
print(r2_summary)

cat("Final Model Performance Summary\n")
cat(sprintf("MAE (Mean Absolute Error): %.2f\n", mae_val))
cat(sprintf("Bayesian R²: %.2f\n", r2_val))

confidence_95 <- with(bayes_data_pred, incidence_numeric >= predicted_lower & incidence_numeric <= predicted_upper)
coverage_rate <- mean(confidence_95, na.rm = TRUE)

ggplot(new_data, aes(x = y_incidence, y = y_pred)) +
  geom_point(color = "blue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(
    title = "Observed vs. Predicted Lung Cancer Incidence",
    x = "Observed Incidence",
    y = "Predicted Incidence"
  ) +
  theme_minimal()

```
